---
title: "Project 2"
author: "Grace Fischer"
date: "10/04/26"
format: 
  html:
    code-overflow: wrap
---

# Setting Up
```{r}
#| label: Set-Up
#| warning: false
#| echo: true

## Installing and calling library to prepare for assignment tasks 

install.packages("tidyverse")
install.packages("tidytuesdayR")
library(tidyverse)
library(tidytuesdayR)

```

# Part 1: Fun with Functions 
## Part 1A: Exponentional Transformation
```{r}
#| label: Exponentional Transformation
#| warning: false 
#| echo: true 

## Pasting function set up from assignment details
Exp <- function(x, k) {
    ## Accounting for the constant term
    result <- 1 
    ## Using i to make seq_len from Lecture 8 Notes to make sure we have as i goes to k 
    for (i in seq_len(k)){
        result <- result + x^i / factorial(i)
    }
    result
}

## Example output: 
cat("First example result from Exp(x = 1, k = 1):", Exp(x = 1, k = 1), "\n") ## This is simply, 1 + 1, output should be 2

cat("Second example result from Exp(x = 4, k = 2):", Exp(x = 4, k = 2), "\n") ## Output should be 13
```


## Part 1B: Sample Mean and Standard Deviation
```{r}
#| label: Sample Mean and Standard Deviation
#| warning: false 
#| echo: true 

sample_mean <- function(x) {
    ## Set length of the vector x to be N for the sample 
    N <- length(x)
    ## Instantiating total variable
    total <- 0 
    ## Using seq_along to loop over a vector this time, like in Lecture 8 notes 
    for (i in seq_along(x)){
        ## Updating total variable 
        total <- total + x[i]
    }
    total / N
}

sample_sd <- function(x) {
 ## Set length of the vector x to be N for the sample
    N <- length(x)
    ## Instantiating ss variable
    ss <- 0 
    ## Creating xbar varaible with sample_mean function we created 
    xbar <- sample_mean(x)
    ## Using seq_along to loop over a vector this time, like in Lecture 8 notes 
    for (i in seq_along(x)){
        ## Updating ss variable using first part of formula
        ss <- ss + (x[i] - xbar)^2
    }
    sqrt(ss/ (N - 1))
}

## Example outputs 
## I based expected outputs based off hand calculations, so I rounded (hence the tilda's)

x1 <- c(1,2,3,4,5,6,7,8,9,10)
cat("From x1 <- c(1,2,3,4,5,6,7,8,9,10), we see a sample mean of", sample_mean(x1), "and a sample sd of", sample_sd(x1), "\n")## Expected output: 5.5, Expected output: ~ 3.02 

x2 <- c(2,4,6,8,8,6,4,2)
cat("From x2 <- c(2,4,6,8,8,6,4,2), we see a sample mean of", sample_mean(x2), "and a sample sd of", sample_sd(x2), "\n")## Expected output: 5, Expected output: ~ 2.4

```


## Part 1C: Confidence Intervas
```{r}
#| label: Confidence Intervals
#| warning: false 
#| echo: true 

## Pasted from assignment directions
calculate_CI <- function(x, conf = 0.95) {
    ## Creating variables based on previous parts of the problem, and standard error calculations of mean 
    N <- length(x)
    xbar <- sample_mean(x)
    se <- sample_sd(x) / sqrt(N) 

    alpha <- 1 - conf
    degrees_freedom <- N - 1
    t_score <- qt(p = alpha / 2, df = degrees_freedom, lower.tail = FALSE)

    margin <- t_score * se
    c(lower_bound = xbar - margin, upper_bound = xbar + margin)
}

## Example outputs
## Using x1, and x2 from earlier where 
## x1 <- c(1,2,3,4,5,6,7,8,9,10) 
## x2 <- c(2,4,6,8,8,6,4,2)

##calculate_CI(x1, conf = 0.95)

cat("Using our function we created to find confidence interval using x1 from ealier and conf = 95, we see",calculate_CI(x1, conf = 0.95), "\n")

## Checking against R function
dat1 <- data.frame(x = x1)
fit1 <- lm(x ~ 1, dat1)
##confint(fit1, level = 0.95)

cat("Using the R function to check out function, we see",confint(fit1, level = 0.95), "\n")

##calculate_CI(x2, conf = 0.98)

cat("Using our function we created to find confidence interval using x2 from ealier and conf = 98, we see", calculate_CI(x2, conf = 0.98), "\n")

## Checking against R function
dat2 <- data.frame(x = x2)
fit2 <- lm(x ~ 1, dat2)
##confint(fit2, level = 0.98)

cat("Using the R function to check out function, we see", confint(fit2, level = 0.98), "\n")
```

# Part 2: Wrangling Data

## Setting Up Data for Part 2
```{r}
#| label: Downloading Part 2 data
#| warning: false 
#| echo: true 

## Copy and pasted from website instructions

tuesdata <- tidytuesdayR::tt_load("2020-01-07")
rainfall <- tuesdata$rainfall
temperature <- tuesdata$temperature

```

```{r}
#| label: Checking for data 
#| warning: false 
#| echo: true 

## Copy and pasted from website instructions
library(here)

## Added directory line for additional repo 
if (!dir.exists(here("data"))) {
    dir.create(here("data"))
}

if (!file.exists(here("data", "tuesdata_rainfall.RDS"))) {
    tuesdata <- tidytuesdayR::tt_load("2020-01-07")
    rainfall <- tuesdata$rainfall
    temperature <- tuesdata$temperature

    # save the files to RDS objects
    saveRDS(tuesdata$rainfall, file = here("data", "tuesdata_rainfall.RDS"))
    saveRDS(tuesdata$temperature, file = here("data", "tuesdata_temperature.RDS"))
}
```

```{r}
#| label: Loading In Data 
#| warning: false 
#| echo: true 

## Copy and pasted from website instructions 

rainfall <- readRDS(here("data", "tuesdata_rainfall.RDS"))
temperature <- readRDS(here("data", "tuesdata_temperature.RDS"))

library(tidyverse)

glimpse(rainfall)

glimpse(temperature)
```

## Task Portion
```{r}
#| label: Creating a new data frame
#| warning: false 
#| echo: true 

## Check dim in rainfall 
dim(rainfall) 

## Creating data frame 
df <- rainfall %>%
    ## Dropping NA's from the columnms (source below)
    drop_na() %>%

    ## remove = false (source below)
    unite(col = "date", year:day, sep = "-",
          remove = FALSE) %>%

    ## ymd from lubricate (source below)
    mutate(date = ymd(date)) %>%
    select(-month, -day) %>%

    ## using str fx from assignment hint (Source below)
    mutate(city_name = str_to_upper(city_name)) %>%
    inner_join(temperature, by = c("city_name", "date"), relationship = "many-to-many")

## Checking new dimmensions of df to make sure edits were made 
dim(df) 

```

Sources I used to help me complete this section: 
https://tidyr.tidyverse.org/reference/drop_na.html
https://tidyr.tidyverse.org/reference/unite.html
https://lubridate.tidyverse.org/reference/ymd.html
https://stringr.tidyverse.org/reference/case.html



# Part 3: Data Visualization
```{r}
#| label: Install color packages 
#| warning: false 
#| echo: true 
install.packages("paletteer")
install.packages("ggthemes")
```

Sources I used to help me complete this: 
https://r-charts.com/color-palettes/

## Part 3A: Plotting temperature data over time
```{r}
#| label: Plotting temperature data over time
#| warning: false 
#| echo: true 

## library for unique color values 
library(paletteer)
library(ggthemes)

## Editing data frame 
df %>%
    filter(year >= 2014) %>%
    
    ## Making plot w aes, setting x and y respectively
    ggplot(aes(x = date, y = temperature, color = temp_type)) +
    geom_line(alpha = 0.9) +
    facet_grid(city_name ~ .) +

    ## Scale taken from palentteer package
    scale_colour_manual(

        ## Hex Codes source below
        values = paletteer_c("ggthemes::Classic Orange-Blue", 30)[c(8, 23)],
        ## If I set it normally, I would do it this way: 
        ## values = c("max" = "orange", "min" = "blue"),
        labels = c("max" = "Daily max", "min" = "Daily min")
    ) +

    ## Setting x and y labels, as well as titles and subtitles
    labs(
        title = "Daily Maximum and Minimum Temperatures in Australian Cities, 2014–2019",
        subtitle = "All cities follow a clear summer–winter cycle; Brisbane is warmest on average,\nwhile Canberra has the widest gap between highs and lows",
        x = "Date",
        y = "Temperature (in Degrees Celsius)",
        color = "Temperature"
    ) +
    theme_bw(base_family = "Times") +

    ## Source below
    theme(
        plot.title = element_text(face = "bold", size = 14),
        legend.position = "bottom"
    )

```

Sources I used to help me complete this section: 
https://ggplot2.tidyverse.org/reference/theme.html
https://r-charts.com/color-palettes/

## Part 3B: Plotting rainfall over time
```{r}
#| label: Plotting rainfall over time
#| warning: false 
#| echo: true 
#| error: true 

## 


```