---
title: "Project 2"
author: "Grace Fischer"
date: "10/04/26"
format: 
  html:
    code-overflow: wrap
---

# Setting Up
```{r}
#| label: Set-Up
#| warning: false
#| eval: false
#| echo: true

## Installing and calling library to prepare for assignment tasks 

install.packages("tidyverse")
install.packages("tidytuesdayR")

```

# Part 1: Fun with Functions 
## Part 1A: Exponentional Transformation
```{r}
#| label: Exponentional Transformation
#| warning: false 
#| echo: true 

## Pasting function set up from assignment details
library(tidyverse)
library(tidytuesdayR)

Exp <- function(x, k) {
    ## Accounting for the constant term
    result <- 1 
    ## Using i to make seq_len from Lecture 8 Notes to make sure we have as i goes to k 
    for (i in seq_len(k)){
        result <- result + x^i / factorial(i)
    }
    result
}

## Example output: 
cat("First example result from Exp(x = 1, k = 1):", Exp(x = 1, k = 1), "\n") ## This is simply, 1 + 1, output should be 2

cat("Second example result from Exp(x = 4, k = 2):", Exp(x = 4, k = 2), "\n") ## Output should be 13
```


## Part 1B: Sample Mean and Standard Deviation
```{r}
#| label: Sample Mean and Standard Deviation
#| warning: false 
#| echo: true 

sample_mean <- function(x) {
    ## Set length of the vector x to be N for the sample 
    N <- length(x)
    ## Instantiating total variable
    total <- 0 
    ## Using seq_along to loop over a vector this time, like in Lecture 8 notes 
    for (i in seq_along(x)){
        ## Updating total variable 
        total <- total + x[i]
    }
    total / N
}

sample_sd <- function(x) {
 ## Set length of the vector x to be N for the sample
    N <- length(x)
    ## Instantiating ss variable
    ss <- 0 
    ## Creating xbar varaible with sample_mean function we created 
    xbar <- sample_mean(x)
    ## Using seq_along to loop over a vector this time, like in Lecture 8 notes 
    for (i in seq_along(x)){
        ## Updating ss variable using first part of formula
        ss <- ss + (x[i] - xbar)^2
    }
    sqrt(ss/ (N - 1))
}

## Example outputs 
## I based expected outputs based off hand calculations, so I rounded (hence the tildes)

x1 <- c(1,2,3,4,5,6,7,8,9,10)
cat("From x1 <- c(1,2,3,4,5,6,7,8,9,10), we see a sample mean of", sample_mean(x1), "and a sample sd of", sample_sd(x1), "\n")## Expected output: 5.5, Expected output: ~ 3.02 

x2 <- c(2,4,6,8,8,6,4,2)
cat("From x2 <- c(2,4,6,8,8,6,4,2), we see a sample mean of", sample_mean(x2), "and a sample sd of", sample_sd(x2), "\n")## Expected output: 5, Expected output: ~ 2.4

```


## Part 1C: Confidence Intervals
```{r}
#| label: Confidence Intervals
#| warning: false 
#| echo: true 

## Pasted from assignment directions
calculate_CI <- function(x, conf = 0.95) {
    ## Creating variables based on previous parts of the problem, and standard error calculations of mean 
    N <- length(x)
    xbar <- sample_mean(x)
    se <- sample_sd(x) / sqrt(N) 

    alpha <- 1 - conf
    degrees_freedom <- N - 1
    t_score <- qt(p = alpha / 2, df = degrees_freedom, lower.tail = FALSE)

    margin <- t_score * se
    c(lower_bound = xbar - margin, upper_bound = xbar + margin)
}

## Example outputs
## Using x1, and x2 from earlier where 
## x1 <- c(1,2,3,4,5,6,7,8,9,10) 
## x2 <- c(2,4,6,8,8,6,4,2)

##calculate_CI(x1, conf = 0.95)

cat("Using our function we created to find confidence interval using x1 from earlier and conf = .95, we see",calculate_CI(x1, conf = 0.95), "\n")

## Checking against R function
dat1 <- data.frame(x = x1)
fit1 <- lm(x ~ 1, dat1)
##confint(fit1, level = 0.95)

cat("Using the R function to check out function, we see",confint(fit1, level = 0.95), "\n")

##calculate_CI(x2, conf = 0.98)

cat("Using our function we created to find confidence interval using x2 from ealier and conf = .98, we see", calculate_CI(x2, conf = 0.98), "\n")

## Checking against R function
dat2 <- data.frame(x = x2)
fit2 <- lm(x ~ 1, dat2)
##confint(fit2, level = 0.98)

cat("Using the R function to check out function, we see", confint(fit2, level = 0.98), "\n")
```

# Part 2: Wrangling Data

## Setting Up Data for Part 2
```{r}
#| label: Downloading Part 2 data
#| warning: false 
#| echo: true 

## Copy and pasted from website instructions

tuesdata <- tidytuesdayR::tt_load("2020-01-07")
rainfall <- tuesdata$rainfall
temperature <- tuesdata$temperature

```

```{r}
#| label: Checking for data 
#| warning: false 
#| echo: true 

## Copy and pasted from website instructions
library(here)

## Added directory line for additional repo 
if (!dir.exists(here("data"))) {
    dir.create(here("data"))
}

if (!file.exists(here("data", "tuesdata_rainfall.RDS"))) {
    tuesdata <- tidytuesdayR::tt_load("2020-01-07")
    rainfall <- tuesdata$rainfall
    temperature <- tuesdata$temperature

    # save the files to RDS objects
    saveRDS(tuesdata$rainfall, file = here("data", "tuesdata_rainfall.RDS"))
    saveRDS(tuesdata$temperature, file = here("data", "tuesdata_temperature.RDS"))
}
```

```{r}
#| label: Loading In Data 
#| warning: false 
#| echo: true 

## Copy and pasted from website instructions 

rainfall <- readRDS(here("data", "tuesdata_rainfall.RDS"))
temperature <- readRDS(here("data", "tuesdata_temperature.RDS"))

library(tidyverse)

glimpse(rainfall)

glimpse(temperature)
```

## Task Portion
```{r}
#| label: Creating a new data frame
#| warning: false 
#| echo: true 

## Check dim in rainfall 
dim(rainfall) 

## Creating data frame 
df <- rainfall %>%
    ## Dropping NA's from the columnms (source below)
    drop_na() %>%

    ## remove = false (source below)
    unite(col = "date", year:day, sep = "-",
          remove = FALSE) %>%

    ## ymd from lubridate (source below)
    mutate(date = ymd(date)) %>%
    select(-month, -day) %>%

    ## using str fx from assignment hint (Source below)
    mutate(city_name = str_to_upper(city_name)) %>%
    inner_join(temperature, by = c("city_name", "date"), relationship = "many-to-many")

## Checking new dimmensions of df to make sure edits were made 
dim(df) 

```

Sources I used to help me complete this section: 

https://tidyr.tidyverse.org/reference/drop_na.html

https://tidyr.tidyverse.org/reference/unite.html

https://lubridate.tidyverse.org/reference/ymd.html

https://stringr.tidyverse.org/reference/case.html



# Part 3: Data Visualization
```{r}
#| label: Install color packages 
#| warning: false 
#| eval: false
#| echo: true 
install.packages("paletteer")
install.packages("ggthemes")
```

Sources I used to help me complete this: 

https://r-charts.com/color-palettes/

## Part 3A: Plotting temperature data over time
```{r}
#| label: Plotting temperature data over time
#| warning: false 
#| echo: true 

## library for unique color values 
library(paletteer)
library(ggthemes)

## Editing data frame 
df %>%
    filter(year >= 2014) %>%
    
    ## Making plot w aes, setting x and y respectively
    ggplot(aes(x = date, y = temperature, color = temp_type)) +
    geom_line(alpha = 0.9) +
    facet_grid(city_name ~ .) +

    ## Scale taken from paletteer package
    scale_colour_manual(

        ## Hex Codes source below
        values = paletteer_c("ggthemes::Classic Orange-Blue", 30)[c(8, 23)],
        ## If I set it normally, I would do it this way: 
        ## values = c("max" = "orange", "min" = "blue"),
        labels = c("max" = "Daily max", "min" = "Daily min")
    ) +

    ## Setting x and y labels, as well as titles and subtitles
    labs(
        x = "Date",
        y = "Temperature (in Degrees Celsius)",
        color = "Temperature",
        title = "Daily Maximum and Minimum Temperatures in Australian Cities, 2014–2019",
        subtitle = "All cities follow a clear summer–winter cycle; Brisbane is warmest on average,\nwhile Canberra has the widest gap between highs and lows",
        caption = "Data: Australian Bureau of Meteorology"
    ) +
    theme_bw(base_family = "Times") +

    ## Source below
    theme(
        plot.title = element_text(face = "bold", size = 14),
        legend.position = "bottom"
    )

```

Sources I used to help me complete this section: 

https://ggplot2.tidyverse.org/reference/theme.html

https://r-charts.com/color-palettes/

## Part 3B: Plotting rainfall over time
```{r}
#| label: Plotting rainfall over time
#| warning: false 
#| echo: true 

## Creating function
plot_rainfall <- function(city_name, year) {
    ## Check inputs so that exactly one city and one year were entered
    if (length(city_name) != 1L || length(year) != 1L) {
        stop("Please enter one city and one year at a time.", call. = FALSE)
    }

    ## Checks input so that a city was entered as text
    if (!is.character(city_name)) {
        stop("`city_name` must be text, such as \"PERTH\".", call. = FALSE)
    }

    ## Checks input so that year was entered as number
    if (!is.numeric(year)) {
        stop("`year` must be a number, such as 2000.", call. = FALSE)
    }

    ## Rename input to not mess with df 
    city_input <- str_to_upper(city_name)
    year_input <- year

    ## Check to see that the input exists in the data (sources below)
    if (!(city_input %in% df$city_name)) {
        stop("'", city_name, "' is not in the data. Choose from: ",
             paste(unique(df$city_name), collapse = ", "), call. = FALSE)
    }

    ## Filtering for one row per rainfall measurement
    new_df <- df %>%
        filter(city_name == city_input, year == year_input, temp_type == "max")

    if (nrow(new_df) == 0) {
        ## Checking the years from rows where the city matches, like boolean conditions 
        years_available <- unique(df$year[df$city_name == city_input])
        stop("No data for ", city_input, " in ", year_input, ". Years available: ",
             min(years_available), " to ", max(years_available),
             " (some years in between may be missing).", call. = FALSE)
    }


    ## Checking for number of days with no rain 
    days_no_rain <- sum(new_df$rainfall == 0)
    ## Checking for rainy days
    rainy <- new_df %>% filter(rainfall > 0)
    
    ## Creating the plot
    rainy %>%
        ## Plotting log transformation for better visibility 
        ggplot(aes(x = log(rainfall))) +

        ## Setting Color for graph aesthetics / Made bins 20 because I got a message to select better bin number then 30 which is preset value 
        geom_histogram(bins = 20, fill = "deepskyblue2", color = "darkslategrey") +
        labs(
            x = "log(Daily Rainfall in mm)",
            y = "Number of days",
            title = paste("Distribution of Daily Rainfall in", city_input, "in", year_input),
            subtitle = paste0(nrow(rainy), " rainy days (median ",
                              round(median(rainy$rainfall), 1), " mm); ",
                              days_no_rain, " dry days not shown on the log scale"),
            caption = "Data: Australian Bureau of Meteorology"
        ) +
        theme_minimal(base_family = "Georgia")+

    theme(
        plot.title = element_text(face = "bold", size = 14)
    )
}

```

See output in following Cell
```{r}
#| label: output example correct 
#| warning: false 
#| echo: true 

plot_rainfall("Perth", 2000)

```

```{r}
#| label: city error 
#| warning: false 
#| echo: true 
#| error: true

## Error where city is not in data (I googled popular cities and chose the first one not included in data)
plot_rainfall("ADELAIDE", 2000)  

```

```{r}
#| label: year-errors
#| warning: false 
#| echo: true 
#| error: true

## Error where city exists, but year does not 
plot_rainfall("CANBERRA", 1850)
```

```{r}
#| label: args-error
#| warning: false 
#| echo: true 
#| error: true

## Error where too many inputs
plot_rainfall("PERTH", c(2000, 2001))
```

Sources I used to help me complete this:

https://www.geeksforgeeks.org/r-language/collapse-function-in-r/

https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/unique

### Function Design Explanation
I designed `plot_rainfall()` to take two arguments, "city_name" and "year". It has one return which is histogram of daily rainfall on a log scale for that combination. I decided to use the log scale for the histogram to have better visibility/ scale. The repeated calling allows the user to engage in data exploration, without needing to copy and paste the code.

Following the error-checking approach from class, the function checks the user's inputs in different stages prior to doing any of the actual work or graphing. First, it confirms that exactly one city (as text) and one year (as a number) were entered. It then checks that the city exists in the data. Finally, the last error checking measure is that the city has measurements for the requested year. My minor in computer science from undergrad taught me that a good error message should not just tell users that something went wrong, but also show them what they can do instead, so they do not keep making the same mistake. This has remained true through all my CS courses, in various programming languages, so I wanted to try to challenge myself and implement that here in R, since that is not something I have done before. Therefore, each stop() message explains how to fix the input, such as listing the valid cities or the range of years available for that city. I also convert the city name to upper case, so "perth" and "Perth" work as well as "PERTH".

Inside the function, I copy the inputs to new names , because when we filter, it would compare the column to itself and keep every row. I also keep only the rows where temp_type == max, since the join in Part 2 matched each rainfall measurement to both a maximum and a minimum temperature, which would otherwise count every day twice.

Finally, when it comes to plotting, I plotted log(rainfall) because the daily rainfall has a very positive skew. The log transformation just makes the shape of the distribution more visible. Because log(0) is undefined, I exclude days with no rain from the histogram and report how many there were in the subtitle of the plot. The title and subtitle are built with paste(), so they update automatically for each city and year. 


# Part 4: Apply functions and plot
## Part 4A: Tasks
```{r}
#| label: Using Part 1 and Part 2 to make a new df 
#| warning: false 
#| echo: true 

## Create new df
rain_df <- df %>%
    ## Filter for 2014 and onwards, additionally, make one row per rainfall measure
    filter(year >= 2014, temp_type == "max") %>%

    ## Create one group per city and year 
    group_by(city_name, year) %>%

    ## Summarize using the functions we created 
    summarize(
        mean_rainfall = sample_mean(rainfall),
        sd_rainfall   = sample_sd(rainfall),
        lower_bound   = calculate_CI(rainfall, conf = 0.95)["lower_bound"],
        upper_bound   = calculate_CI(rainfall, conf = 0.95)["upper_bound"],
    ) %>%
    ## Do this to avoid error later 
    ungroup()

rain_df

```

## Part 4B: Tasks
```{r}
#| label: Plot using rain_df 
#| warning: false 
#| echo: true 

## Use ggplot and new df to make graph
rain_df %>%

    ## Using aes, setting x and y respectively 
    ggplot(aes(x = year, y = mean_rainfall)) +

    ## Error bars for the 95% confidence interval, sources below 
    geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound),
                  width = 0.2, color = "darkblue") +
    ## Line and points for the means
    geom_line(color = "darkblue") +
    geom_point(color = "darkblue", size = 2) +
    ## Source below
    facet_wrap(~ city_name, nrow = 1) +
    ## ggplot2 function which controls numeric x axis (inspired by Lecture 6 notes, source below)
    scale_x_continuous(breaks = 2014:2019) +
    labs(
        title = "Mean Daily Rainfall in Australian Cities, 2014–2019",
        subtitle = "Perth has the highest mean rainfall per recorded day but the widest intervals, implying there are fewer measurements.\nCanberra is consistently the lowest.\n2019 covers January to May only.",
        caption = "Data: Australian Bureau of Meteorology",
        x = "Year",
        y = "Mean daily rainfall (mm)"
    ) +
    theme_clean(base_family = "Times") +

    theme(
        plot.title = element_text(face = "bold", size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1)
    )
    
    

```

Sources that helped me complete this (from instructions included):

https://r-graphics.org/RECIPE-ANNOTATE-ERROR-BAR.html

https://ggplot2.tidyverse.org/reference/geom_linerange.html

https://ggplot2.tidyverse.org/reference/facet_wrap.html

https://ggplot2.tidyverse.org/reference/scale_continuous.html 
